\textbf{Best practices for applying machine learning to bacterial 16S rRNA gene sequencing data}

### 2019-08-02 submission/manuscript.tex
changes before meeting 2 with Pat

Advances in sequencing technology and decreases in the costs of
generating 16S rRNA gene sequences have allowed rapid exploration of the
human microbiome and its health implications. Currently, the human
microbiome field is growing at an unprecedented rate and there is an
increasing demand for methods that identify associations between
microbiome members and human health. However, this is difficult as human
associated microbial communities are remarkably complex,
high-dimensional and uneven within and between individuals with the same
disease. It is unlikely that a single species can explain a disease.
Instead, subsets of those communities, in relation to one another and to
their host, account for differences in health outcomes.

Machine learning (ML) methods are effective at recognizing and
highlighting patterns in complex microbial datasets. They learn from
existing data to predict the outcomes of new data and allow us to infer
on the reasons underlying that prediction. Therefore, researchers have
started to explore the utility of ML models that use microbiota
associated biomarkers to predict human health and to understand the
ecological basis of diseases such as liver cirrhosis, colorectal cancer,
inflammatory bowel diseases (IBD), obesity, and type 2 diabetes (1--11).
The high-stake task of predicting the correct diagnosis with high
confidence relies on a ML model that is built with rigorous methods.
However, the field's use of ML lacks clarity and consistency in which
methods are used and how these methods are implemented (12, 13). More
notably, we commonly see flawed ML practices such as using ML pipelines
where there is no seperate held-out test dataset to evaluate predictive
performance or reporting few or only the best outcomes of
cross-validation. Even when there are seperate testing sets to evaluate
predictive performance, there are large differences between
cross-validation and testing performances and large confidence intervals
in testing performances (4, 14--20). These practices prevent the
development of generalizable models, where the model makes accurate
predictions with newly acquired data as well as it does with the
training data.

Moreover, there is a lack of discussion on why a particular ML model is
utilized and a lack of emphasis on the strengths and weaknesses of it.
Recently, there is a trend towards using more complex ML models such as
random forest, extreme gradient boosting (XGBoost) and neural networks
without a discussion on if and how much model interpretability is
necessary for the study (11, 21--23). These models are also called black
box ML models because they are not inherently interpretable and require
posthoc explanations to determine the feature importances in making a
prediction. These explanations can be misleading and at times unreliable
when making high-stake decisions about patient health (24). The models
we develop for healthcare, both to predict disease and to understand
underlying reasons behind that prediction, should be transparent and
accountable (25).

The lack of transparency on model selection and interpretation as well
as flawed modeling methods negatively impact model validity and
reproducibility. We need to strive toward better machine learning
practices by (1) implementing rigorous machine learning pipelines and
(2) selecting ML models that reflect the goal of the study as it will
inform our expectations of predictive performance, complexity,
interpretability and computational efficiency. To showcase a rigorous ML
pipeline and to shed light on how ML model selection can affect modeling
results, we performed an empirical analysis comparing several different
ML models using the same dataset and the same ML pipeline. We used a
previously published colorectal cancer (CRC) study (3) which had fecal
16S rRNA gene sequences of CRC patients. The human microbiome is
hypothesized to directly contribute to the development of CRC and fecal
16S rRNA gene sequences have been used to detect it (1, 3, 4, 26). We
built seven ML models using fecal 16S rRNA gene sequences to predict
healthy patients versus patients with colorectal lesions that were
identified by colonoscopy as screen relevant neoplasias (SRN). We
established modeling pipelines for three linear models with different
forms of regularization; L2-regularized logistic regression, L1 and
L2-regularized support vector machines (SVM) with linear kernel. We also
developed four non-linear models; SVM with radial basis function kernel,
a decision tree, random forest and XGBoost. We compared the predictive
performance, interpretability and computational efficiency of the seven
models to highlight the importance of model selection. We established
standards for ML pipeline construction, predictive performance
evaluation and model interpretation for microbiome-associated ML models
that have a binary prediction task.

@@

\textbf{Model selection and pipeline construction}

We first determined the dataset and the ML models to use in our study.
We used a previously published study on a CRC cohort of 490 patients
with 261 cases of SRN. For each patient, we had 6920 features (fecal
bacterial abundances) and a two-class label that defines their
colorectal health (normal or SRN colorectal lesions as defined by
colonoscopies). We established modeling pipelines for a binary
prediction task with L2-regularized logistic regression, L1 and L2
support vector machines (SVM) with linear and radial basis function
kernels, a decision tree, random forest and extreme gradient boosted
decision tree (XGBoost).

We used ML models with different classification algorithms and
regularization parameters. For regularized logistic regression and SVM
with linear kernel, we used L2 regularization to keep all potentially
important features. For comparison, we also trained an L1 regularized
SVM model with linear kernel. L1-regularization on microbiome data lead
to a sparser solution (i.e., force many coefficients to zero). Finally,
to explore the potential for non-linear relationships among features and
the outcome of interest, we trained tree based models, decision tree,
random forest and XGBoost, as well as an SVM with non-linear kernel.

We established a ML pipeline where we train and validate each of the
seven models {[}Figure 1{]}. We randomly split the data into
training/validation and test sets so that the training/validation set
consisted of 80\% of the full dataset while the test set was composed of
the remaining data {[}Figure 1{]}. Since the cases are not uniformly
represented in the data, the initial data-split was stratified to
maintain the overall label distribution in both the training/validation
and test sets. Training/validation set consisted of 393 patients (209
SRN), while the test set was composed of 97 patients (52 SRN). The
training/validation data was used for training purposes and validation
of hyperparameter selection, and the test set was used for evaluation
purposes.

Hyperparameters are the rules that are learned from the training data in
a classification algorithm. When they are tuned over a full grid search
and selected by validation, they make the ML model better are
predicting. We selected hyperparameter settings using repeated five-fold
cross-validation on the training/validation set {[}Figure 1{]}. We chose
the best hyperparameter setting for each model based on its predictive
performance on the validation set using the area under the receiver
operating characteristic curve (AUROC) metric {[}Figure S1 and S2{]}.
The AUROC ranges from 1.0, where the model perfectly distinguishes
between cases and controls, to 0.50, where the model's predictions are
no different from random chance. Similar to the initial data-split,
five-fold cross-validation for hyperparameter selection was also
stratified to maintain the overall label distribution on the training
and validation sets. The cross-validation of each hyperparameter setting
was repeated over 100 randomizations to get a robust reading of
predictive performance.

We then trained the full training/validation dataset with the selected
hyperparameters. We used the held-out test set to evaluate the
predictive performance of each ML model. The data-split, hyperparameter
selection, training and testing steps were repeated 100 times to get a
reliable and robust reading of model performance {[}Figure 1{]}.

\textbf{Predictive performance and generalizability of the seven
models.}

We evaluated the predictive performances of seven binary classification
models when applied to held-out test data using the AUROC metric
{[}Figure 2{]}. Random forest had significantly higher test AUROC values
than the other models for detecting SRNs when AUROC values were compared
to the other six by Wilcoxon rank sum test (p \textless{} 0.01). The
median AUROC of the random forest model was 0.695 (IQR 0.044).
L2-regularized logistic regression, XGBoost, L2-regularized SVM with
linear and radial basis function kernel AUROC values were not
significantly different from one another. They had median AUROC values
of 0.68 (IQR 0.055), 0.679 (IQR 0.052), 0.678 (IQR 0.056) and 0.668 (IQR
0.056) respectively. L1-regularized SVM with linear kernel and decision
tree had significantly lower AUROC values than the other ML models with
median AUROC of 0.65 (IQR 0.066) and 0.601 (IQR 0.059), respectively
{[}Figure 2{]}. Random forest had the highest median AUROC for detecting
SRN. Despite its simplicity, the L2-regularized logistic regression was
second best in predictive performance.

@@

\textbf{Interpretation of each ML model.}

Interpretability is the degree to which humans can understand the
reasons behind a model prediction (27). Because we often use ML models
not just to predict a health outcome but also to learn the ecology
behind a disease, model interpretation becomes crucial for microbiome
studies. The ML models we built using L2-regularized logistic
regression, L1 and L2 support vector machines (SVM) with linear and
radial basis function kernels, a decision tree, random forest and
XGBoost decrease in interpretability as they increase in complexity. In
this study we highlighted two methods to interpret models with varying
complexity.

@@

Microbiome studies use ML models, often with a classification task, to
predict a disease but also to learn which microbes are indicators of
that disease (2--11). Achieving either of these tasks have far-reaching
impact on human health, however ML as a tool in microbiome studies is
still at its infancy. A framework is needed to develop rigorous ML
models, to identify and overcome potential pitfalls. Previous studies
generated workflows to allow ML to be widely used by the microbiome
researchers (28--31). This study sets-up standards for developing and
evaluating rigorous ML models for microbiome data {[}Table 1{]}.

@@

In this study, we also established a rigorous ML pipeline to use 16S
rRNA sequence counts to predict a binary health outcome. First of all,
we used a held-out test set to illustrate the difference between
cross-validation and testing AUROC values. When the difference between
cross-validation and test performance is low, this suggest the models
are not overfit and that they will perform similar with similar data. In
all the models, the difference between median cross-validation and
testing AUROC values did not exceed 0.021 which suggests that these
models are generalizable and can be used to test similar new data.
Furthermore, we performed the initial 80\%-20\% random datasplit 100
times in our ML pipeline. Depending on how the data is split, there is
the chance of being overoptimistic about the predictive performance of a
model. We showed that there was variability in AUROC values between
different random data-splits in each of the models we tested. Our
results showed that the testing AUROC values varied 0.23 on average
between different data-splits. The randomization and resampling of the
initial data-split to create a held-out test set is a crucial step in
the ML pipeline to develop robust ML models and to report reliable
performance metrics. We also performed a full grid search for
hyperparameter settings when training our ML models. Default
hyperparameter settings in previously developed ML packages in R,
Python, and Matlab programming languages are inadequate for effective
application of classification algorithms and need to be optimized for
each new dataset used to generate a model. In the example of
L1-regularized SVM with linear kernel {[}Figure S1{]}, the model showed
large variability between different regularization coefficients (C) and
was susceptible to performing poorly if the wrong regularization
coefficient was assigned to the model by default. And finally, we used
the AUROC metric in our study to evaluate the predictive performance of
the ML models. AUROC is always random at the value 0.5 and is a robust
metric when a dataset is imbalanced.

@@

The colorectal health of the patient was defined as two encompassing
classes; Normal or Screen Relevant Neoplasias (SRNs). Normal class
includes patients with non-advanced adenomas or normal colons whereas
SRN class includes patients with advanced adenomas or carcinomas. The
study had 261 normal and 229 SRN samples. The bacterial abundances are
the features used to predict colonic health of the patients. Bacterial
abundances are discrete data in the form of Operational Taxonomic Unit
(OTU) counts. OTU counts were set to the size of our smallest sample and
were subsampled at the same distances. They were then transformed by
scaling to a {[}0-1{]} range.

@@

\textbf{Figure 5. Computational efficiency of seven ML models.} The
training times for of each data-split showed the differences in
computational efficiency of the seven models. The median training time
in hours was the highest for XGBoost and shortest for L1-regularized SVM
with linear kernel. The boxplot shows quartiles at the box ends and the
statistical median as the horizontal line in the box. The whiskers show
the farthest points that are not outliers. Outliers are data points that
are not within 3/2 times the interquartile ranges. Abbreviations: AUROC,
area under the receiver operating characteristic curve; SVM, support
vector machine; XGBoost, extreme gradient boosting.\\
\newpage
\includegraphics{Figure_S1.png}

@@

\hypertarget{ref-dai_multi-cohort_2018}{}
15. \textbf{Dai Z}, \textbf{Coker OO}, \textbf{Nakatsu G}, \textbf{Wu
WKK}, \textbf{Zhao L}, \textbf{Chen Z}, \textbf{Chan FKL},
\textbf{Kristiansen K}, \textbf{Sung JJY}, \textbf{Wong SH}, \textbf{Yu
J}. 2018. Multi-cohort analysis of colorectal cancer metagenome
identified altered bacteria across populations and universal bacterial
markers. Microbiome \textbf{6}:70.
doi:\href{https://doi.org/10.1186/s40168-018-0451-2}{10.1186/s40168-018-0451-2}.

\hypertarget{ref-montassier_pretreatment_2016}{}
16. \textbf{Montassier E}, \textbf{Al-Ghalith GA}, \textbf{Ward T},
\textbf{Corvec S}, \textbf{Gastinne T}, \textbf{Potel G}, \textbf{Moreau
P}, \textbf{Cochetiere MF de la}, \textbf{Batard E}, \textbf{Knights D}.
2016. Pretreatment gut microbiome predicts chemotherapy-related
bloodstream infection. Genome Medicine \textbf{8}:49.
doi:\href{https://doi.org/10.1186/s13073-016-0301-4}{10.1186/s13073-016-0301-4}.

\hypertarget{ref-papa_non-invasive_2012}{}
17. \textbf{Papa E}, \textbf{Docktor M}, \textbf{Smillie C},
\textbf{Weber S}, \textbf{Preheim SP}, \textbf{Gevers D},
\textbf{Giannoukos G}, \textbf{Ciulla D}, \textbf{Tabbaa D},
\textbf{Ingram J}, \textbf{Schauer DB}, \textbf{Ward DV},
\textbf{Korzenik JR}, \textbf{Xavier RJ}, \textbf{Bousvaros A},
\textbf{Alm EJ}. 2012. Non-invasive mapping of the gastrointestinal
microbiota identifies children with inflammatory bowel disease. PLOS ONE
\textbf{7}:e39242.
doi:\href{https://doi.org/10.1371/journal.pone.0039242}{10.1371/journal.pone.0039242}.

@@

\hypertarget{ref-ai_systematic_2017}{}
19. \textbf{Ai L}, \textbf{Tian H}, \textbf{Chen Z}, \textbf{Chen H},
\textbf{Xu J}, \textbf{Fang J-Y}. 2017. Systematic evaluation of
supervised classifiers for fecal microbiota-based prediction of
colorectal cancer. Oncotarget \textbf{8}:9546--9556.
doi:\href{https://doi.org/10.18632/oncotarget.14488}{10.18632/oncotarget.14488}.

\hypertarget{ref-wong_quantitation_2017}{}
20. \textbf{Wong SH}, \textbf{Kwong TNY}, \textbf{Chow T-C}, \textbf{Luk
AKC}, \textbf{Dai RZW}, \textbf{Nakatsu G}, \textbf{Lam TYT},
\textbf{Zhang L}, \textbf{Wu JCY}, \textbf{Chan FKL}, \textbf{Ng SSM},
\textbf{Wong MCS}, \textbf{Ng SC}, \textbf{Wu WKK}, \textbf{Yu J},
\textbf{Sung JJY}. 2017. Quantitation of faecal fusobacterium improves
faecal immunochemical test in detecting advanced colorectal neoplasia.
Gut \textbf{66}:1441--1448.
doi:\href{https://doi.org/10.1136/gutjnl-2016-312766}{10.1136/gutjnl-2016-312766}.

\hypertarget{ref-galkin_human_2018}{}
21. \textbf{Galkin F}, \textbf{Aliper A}, \textbf{Putin E},
\textbf{Kuznetsov I}, \textbf{Gladyshev VN}, \textbf{Zhavoronkov A}.
2018. Human microbiome aging clocks based on deep learning and tandem of
permutation feature importance and accumulated local effects. bioRxiv.
doi:\href{https://doi.org/10.1101/507780}{10.1101/507780}.

\hypertarget{ref-reiman_using_2017}{}
22. \textbf{Reiman D}, \textbf{Metwally A}, \textbf{Dai Y}. 2017. Using
convolutional neural networks to explore the microbiome, pp. 4269--4272.
\emph{In} 2017 39th annual international conference of the IEEE
engineering in medicine and biology society (EMBC).

\hypertarget{ref-fioravanti_phylogenetic_2017}{}
23. \textbf{Fioravanti D}, \textbf{Giarratano Y}, \textbf{Maggio V},
\textbf{Agostinelli C}, \textbf{Chierici M}, \textbf{Jurman G},
\textbf{Furlanello C}. 2017. Phylogenetic convolutional neural networks
in metagenomics. arXiv:170902268 {[}cs, q-bio{]}.

\hypertarget{ref-rudin_please_2018}{}
24. \textbf{Rudin C}. 2018. Please stop explaining black box models for
high stakes decisions. arXiv:181110154 {[}cs, stat{]}.

\hypertarget{ref-rudin_optimized_2018}{}
25. \textbf{Rudin C}, \textbf{Ustun B}. 2018. Optimized scoring systems:
Toward trust in machine learning for healthcare and criminal justice.
Interfaces \textbf{48}:449--466.
doi:\href{https://doi.org/10.1287/inte.2018.0957}{10.1287/inte.2018.0957}.

\hypertarget{ref-knights_human-associated_2011}{}
26. \textbf{Knights D}, \textbf{Parfrey LW}, \textbf{Zaneveld J},
\textbf{Lozupone C}, \textbf{Knight R}. 2011. Human-associated microbial
signatures: Examining their predictive value. Cell Host Microbe
\textbf{10}:292--296.
doi:\href{https://doi.org/10.1016/j.chom.2011.09.003}{10.1016/j.chom.2011.09.003}.

\hypertarget{ref-miller_explanation_2017}{}
27. \textbf{Miller T}. 2017. Explanation in artificial intelligence:
Insights from the social sciences. arXiv:170607269 {[}cs{]}.

\hypertarget{ref-statnikov_comprehensive_2013}{}
28. \textbf{Statnikov A}, \textbf{Henaff M}, \textbf{Narendra V},
\textbf{Konganti K}, \textbf{Li Z}, \textbf{Yang L}, \textbf{Pei Z},
\textbf{Blaser MJ}, \textbf{Aliferis CF}, \textbf{Alekseyenko AV}. 2013.
A comprehensive evaluation of multicategory classification methods for
microbiomic data. Microbiome \textbf{1}:11.
doi:\href{https://doi.org/10.1186/2049-2618-1-11}{10.1186/2049-2618-1-11}.

\hypertarget{ref-knights_supervised_2011}{}
29. \textbf{Knights D}, \textbf{Costello EK}, \textbf{Knight R}. 2011.
Supervised classification of human microbiota. FEMS Microbiology Reviews
\textbf{35}:343--359.
doi:\href{https://doi.org/10.1111/j.1574-6976.2010.00251.x}{10.1111/j.1574-6976.2010.00251.x}.

>>>>

As the number of people represented in human microbiome datasets grow,
there is an increasing desire to use microbiome data to diagnose
disease. However, the structure of the human microbiome is remarkably
variable between individuals to the point where it is often difficult to
identify the bacterial populations that are associated with diseases
using traditional statistical models. This variation is likely due to
the ability of many bacterial populations to fill the same niche such
that different populations cause the same disease in different
individuals. Furthermore, a growing number of studies have shown that it
is rare for a single bacterial species to be associated with a disease.
Instead, subsets of the microbiome account for differences in health.
Traditional statistical approaches do not adequately account for the
variation in the human microbiome and typically consider the protective
or risk effects of each bacterial population individually. Recently,
machine learning models have grown in popularity among microbiome
researchers because of the large amount of data that can now be
generated and because the models are effective at accounting for the
interpersonal microbiome variation and the ecology of the disease.

ML models are useful for understanding the variation in the structure of
existing data and to apply that knowledge to make predictions about new
data. Researchers have used ML models to diagnose and understand the
ecological basis of diseases such as liver cirrhosis, colorectal cancer,
inflammatory bowel diseases (IBD), obesity, and type 2 diabetes (1--11).
The task of diagnosing an individual with high confidence relies on a ML
model that is built with rigorous methods. However, ML methodology in
the microbiome field has flaws that have not been addressed previously.
There is a lack of transparency in which methods are used and how these
methods are implemented (12, 13); models are being developed and
evaluated without a seperate held-out test data (3, 4, 14--16); there is
large variation between the cross-validation outcomes (16, 17) and
between cross-validation and testing performances (18) . These practices
limit reproducibility, cause overoptimism for model performance and
prevent the development of generalizable models, where the model makes
accurate predictions with newly acquired data as well as it does with
the training data. Nevertheless, the microbiome field is making progress
to avoid potential pitfalls of ML. More and more studies are now
validating their models on independent datasets (7, 18, 19) and they are
introducing frameworks to accurately use ML tools (20--23).

Among microbiome researchers, there has been a trend towards using more
complex ML models such as random forest and neural networks (2, 11,
24--26) over simpler models such as logistic regression or other linear
models (18, 22, 27). There is an implicit assumption that the more
complex models are better because they are more complex. Although these
models may be better at incorporating non-linear relationships or yield
better predictions, they are also called black box ML models because
they are not inherently interpretable. These models require post hoc
explanations to quantify the importance of each feature in making a
prediction. Depending on the application of the model researchers may
chose to use different modeling approaches. For example, a researcher
trying to identify the populations causing a disease would likely want a
more interpretable model whereas a clinician may emphasize performance.
Although one may feel that they are sacrificing interpretability for
performance, that tradeoff may be minimal (28, 29).

The lack of transparency in model selection and using flawed modelling
and interpretation methods negatively impacts model validity. To
showcase a rigorous ML pipeline and to shed light on how ML model
selection can affect modeling results, we performed an empirical
analysis comparing 7 modeling approaches using the same dataset and
pipeline. We established modeling pipelines for three linear models with
different forms of regularization: L2-regularized logistic regression
and L1 and L2-regularized support vector machines (SVM) with a linear
kernel. We also developed four non-linear models: SVM with radial basis
function kernel, a decision tree, random forest and XGBoost. We compared
the predictive performance, interpretability and computational
efficiency of the seven models to highlight the importance of model
selection. To demonstrate the performance of these modelling approaches
and our pipeline, we used data from a previously published study that
sought to classifiy individuals as having normal colons or colonic
lesions based on the 16S rRNA gene sequences collected from fecal
samples (3). This dataset was selected because it is a relatively large
collection of samples (N=490) connected to clinically significant
disease where there is ample evidence that it is driven by variation in
the microbiome (1, 3, 4, 30). With this dataset, we established
standards for ML pipeline construction, evaluated predictive
performance, and demonstrated model interpretation for models. This
framework can be easily applied to other host-associated and
environmental microbiome problems.

@@

\textbf{Model selection and pipeline construction} We established a ML
pipeline where we train and validate each of the seven models {[}Figure
1{]}. We randomly split the data into training and test sets so that the
training set consisted of 80\% of the full dataset while the test set
was composed of the remaining data {[}Figure 1{]}. Since the cases are
not uniformly represented in the data, the data-split was stratified to
maintain the overall label distribution in both the training and test
sets. For example, our example dataset represented data from 490
individuals, 261 had normal colons and 229 had a screen relevant
neoplasia (SRN). After the data-split, the training set consisted of 393
patients (209 SRN), while the test set was composed of 97 patients (52
SRN). The training data was used for hyperparameter selection and the
test set was used for evaluating predictive performance.

We used ML models with different classification algorithms and
regularization methods. Regularization is a technique to discourage
overfitting by penalizing the model for learning the training data too
well. For regularized logistic regression and SVM with linear kernel, we
used L2 regularization to keep all potentially important features. For
comparison, we also trained an L1 regularized SVM model with linear
kernel. L1-regularization on microbiome data lead to a sparser solution
(i.e., force many coefficients to zero). Finally, to explore the
potential for non-linear relationships among features and the outcome of
interest, we trained tree based models, decision tree, random forest and
XGBoost, as well as an SVM with non-linear kernel.

Hyperparameters are the rules that are learned from the training data in
a classification algorithm. For example regularization term (C) that
decides how big the penalty for overfitting will be, is an
hyperparameter. The model becomes more generalizable when we tune for
the optimal C value. Similar to regularization term C, other
hyperparameters can be tuned over a full grid search and selected by
validation to build better models. We selected hyperparameter settings
by performing repeated five-fold cross-validation (CV) on the training
set {[}Figure 1{]}. Similar to the initial data-split, five-fold CV was
also stratified to maintain the overall label distribution. We chose the
best hyperparameter setting for each model based on its CV predictive
performance using the area under the receiver operating characteristic
curve (AUROC) metric {[}Figure S1 and S2{]}. The AUROC ranges from 1.0,
where the model perfectly distinguishes between cases and controls, to
0.50, where the model's predictions are no different from random chance.
The cross-validation of each hyperparameter setting was repeated over
100 randomizations to get a robust reading of predictive performance.

We then trained the full training dataset with the selected
hyperparameters. We used the held-out test set to evaluate the testing
predictive performance of each ML model. The data-split, hyperparameter
selection, training and testing steps were repeated 100 times to get a
reliable and robust reading of model performance {[}Figure 1{]}.

\textbf{Predictive performance and generalizability of the seven
models.} We evaluated the predictive performances of seven binary
classification models when applied to held-out test data using the AUROC
metric {[}Figure 2{]}. Random forest had significantly higher test AUROC
values than the other models for detecting SRNs when AUROC values were
compared to the other six by Wilcoxon rank sum test (p \textless{}
0.01). The median AUROC of the random forest model was 0.695 (IQR
0.044). L2-regularized logistic regression, XGBoost, L2-regularized SVM
with linear and radial basis function kernel AUROC values were not
significantly different from one another. They had median AUROC values
of 0.68 (IQR 0.055), 0.679 (IQR 0.052), 0.678 (IQR 0.056) and 0.668 (IQR
0.056) respectively. L1-regularized SVM with linear kernel and decision
tree had significantly lower AUROC values than the other ML models with
median AUROC of 0.65 (IQR 0.066) and 0.601 (IQR 0.059), respectively
{[}Figure 2{]}. Random forest had the highest median AUROC for detecting
SRN. Despite its simplicity, the L2-regularized logistic regression was
second best in predictive performance.

@@

\textbf{Interpretation of each ML model.} Interpretability is the degree
to which humans can understand the reasons behind a model prediction
(31). Because we often use ML models not just to predict a health
outcome but also to learn the ecology behind a disease, model
interpretation becomes crucial for microbiome studies. The ML models we
built using L2-regularized logistic regression, L1 and L2 support vector
machines (SVM) with linear and radial basis function kernels, a decision
tree, random forest and XGBoost decrease in interpretability as they
increase in complexity. In this study we highlighted two methods to
interpret models with varying complexity.

@@

Microbiome studies use ML models, often with a classification task, to
predict a disease but also to learn which microbes are indicators of
that disease (2--11). Achieving either of these tasks have far-reaching
impact on human health, however ML as a tool in microbiome studies is
still at its infancy. A framework is needed to develop rigorous ML
models, to identify and overcome potential pitfalls. Previous studies
generated workflows to allow ML to be widely used by the microbiome
researchers (20--23). This study sets-up standards for developing and
evaluating rigorous ML models for microbiome data {[}Table 1{]}.

@@

In this study, we also established a rigorous ML pipeline to use 16S
rRNA sequence counts to predict a binary health outcome. First of all,
we used a held-out test set to illustrate the difference between
cross-validation and testing AUROC values. When the difference between
cross-validation and test performance is low, this suggest the models
are not overfit and that they will perform similar with similar data. In
all the models, the difference between median cross-validation and
testing AUROC values did not exceed 0.021 which suggests that these
models are generalizable and can be used to test similar new data.
Furthermore, we performed the initial 80\%-20\% random datasplit 100
times in our ML pipeline. Depending on how the data is split, there is
the chance of being overoptimistic about the predictive performance of a
model. We showed that there was variability in AUROC values between
different random data-splits in each of the models we tested. Our
results showed that the testing AUROC values varied 0.23 on average
between different data-splits. The randomization and resampling of the
initial data-split to create a held-out test set is a crucial step in
the ML pipeline to develop robust ML models and to report reliable
performance metrics. Additionally, we performed a full grid search for
hyperparameter settings when training our ML models. Default
hyperparameter settings in previously developed ML packages in R,
Python, and Matlab programming languages are inadequate for effective
application of classification algorithms and need to be optimized for
each new dataset used to generate a model. In the example of
L1-regularized SVM with linear kernel {[}Figure S1{]}, the model showed
large variability between different regularization coefficients (C) and
was susceptible to performing poorly if the wrong regularization
coefficient was assigned to the model by default. And finally, we used
the AUROC metric in our study to evaluate the predictive performance of
the ML models. AUROC is always random at the value 0.5 and is a robust
metric when a dataset is imbalanced.

@@

The colorectal health of the patient was defined as two encompassing
classes; Normal or Screen Relevant Neoplasias (SRNs). Normal class
includes patients with non-advanced adenomas or normal colons whereas
SRN class includes patients with advanced adenomas or carcinomas. The
study had 261 normal and 229 SRN samples. The bacterial abundances are
the features used to predict colorectal health of the patients. For each
patient, we had 6920 features (fecal bacterial abundances) and a
two-class label that defines their colorectal health (normal or SRN
colorectal lesions as defined by colonoscopies). We established modeling
pipelines for a binary prediction task Bacterial abundances are discrete
data in the form of Operational Taxonomic Unit (OTU) counts. OTU counts
were set to the size of our smallest sample and were subsampled at the
same distances. They were then transformed by scaling to a {[}0-1{]}
range.

@@

\textbf{Figure 5. Computational efficiency of seven ML models.} The
training times for of each data-split showed the differences in
computational efficiency of the seven models. The median training time
in hours was the highest for XGBoost and shortest for L1-regularized SVM
with linear kernel. The boxplot shows quartiles at the box ends and the
statistical median as the horizontal line in the box. The whiskers show
the farthest points that are not outliers. Outliers are data points that
are not within 3/2 times the interquartile ranges. Abbreviations: AUROC,
area under the receiver operating characteristic curve; SVM, support
vector machine; XGBoost, extreme gradient boosting. \newpage
\includegraphics{Figure_S1.png}

@@

\hypertarget{ref-montassier_pretreatment_2016}{}
15. \textbf{Montassier E}, \textbf{Al-Ghalith GA}, \textbf{Ward T},
\textbf{Corvec S}, \textbf{Gastinne T}, \textbf{Potel G}, \textbf{Moreau
P}, \textbf{Cochetiere MF de la}, \textbf{Batard E}, \textbf{Knights D}.
2016. Pretreatment gut microbiome predicts chemotherapy-related
bloodstream infection. Genome Medicine \textbf{8}:49.
doi:\href{https://doi.org/10.1186/s13073-016-0301-4}{10.1186/s13073-016-0301-4}.

\hypertarget{ref-ai_systematic_2017}{}
16. \textbf{Ai L}, \textbf{Tian H}, \textbf{Chen Z}, \textbf{Chen H},
\textbf{Xu J}, \textbf{Fang J-Y}. 2017. Systematic evaluation of
supervised classifiers for fecal microbiota-based prediction of
colorectal cancer. Oncotarget \textbf{8}:9546--9556.
doi:\href{https://doi.org/10.18632/oncotarget.14488}{10.18632/oncotarget.14488}.

\hypertarget{ref-dai_multi-cohort_2018}{}
17. \textbf{Dai Z}, \textbf{Coker OO}, \textbf{Nakatsu G}, \textbf{Wu
WKK}, \textbf{Zhao L}, \textbf{Chen Z}, \textbf{Chan FKL},
\textbf{Kristiansen K}, \textbf{Sung JJY}, \textbf{Wong SH}, \textbf{Yu
J}. 2018. Multi-cohort analysis of colorectal cancer metagenome
identified altered bacteria across populations and universal bacterial
markers. Microbiome \textbf{6}:70.
doi:\href{https://doi.org/10.1186/s40168-018-0451-2}{10.1186/s40168-018-0451-2}.

@@

\hypertarget{ref-wong_quantitation_2017}{}
19. \textbf{Wong SH}, \textbf{Kwong TNY}, \textbf{Chow T-C}, \textbf{Luk
AKC}, \textbf{Dai RZW}, \textbf{Nakatsu G}, \textbf{Lam TYT},
\textbf{Zhang L}, \textbf{Wu JCY}, \textbf{Chan FKL}, \textbf{Ng SSM},
\textbf{Wong MCS}, \textbf{Ng SC}, \textbf{Wu WKK}, \textbf{Yu J},
\textbf{Sung JJY}. 2017. Quantitation of faecal fusobacterium improves
faecal immunochemical test in detecting advanced colorectal neoplasia.
Gut \textbf{66}:1441--1448.
doi:\href{https://doi.org/10.1136/gutjnl-2016-312766}{10.1136/gutjnl-2016-312766}.

\hypertarget{ref-statnikov_comprehensive_2013}{}
20. \textbf{Statnikov A}, \textbf{Henaff M}, \textbf{Narendra V},
\textbf{Konganti K}, \textbf{Li Z}, \textbf{Yang L}, \textbf{Pei Z},
\textbf{Blaser MJ}, \textbf{Aliferis CF}, \textbf{Alekseyenko AV}. 2013.
A comprehensive evaluation of multicategory classification methods for
microbiomic data. Microbiome \textbf{1}:11.
doi:\href{https://doi.org/10.1186/2049-2618-1-11}{10.1186/2049-2618-1-11}.

\hypertarget{ref-knights_supervised_2011}{}
21. \textbf{Knights D}, \textbf{Costello EK}, \textbf{Knight R}. 2011.
Supervised classification of human microbiota. FEMS Microbiology Reviews
\textbf{35}:343--359.
doi:\href{https://doi.org/10.1111/j.1574-6976.2010.00251.x}{10.1111/j.1574-6976.2010.00251.x}.

\hypertarget{ref-wirbel_meta-analysis_2019}{}
22. \textbf{Wirbel J}, \textbf{Pyl PT}, \textbf{Kartal E}, \textbf{Zych
K}, \textbf{Kashani A}, \textbf{Milanese A}, \textbf{Fleck JS},
\textbf{Voigt AY}, \textbf{Palleja A}, \textbf{Ponnudurai R},
\textbf{Sunagawa S}, \textbf{Coelho LP}, \textbf{Schrotz-King P},
\textbf{Vogtmann E}, \textbf{Habermann N}, \textbf{Niméus E},
\textbf{Thomas AM}, \textbf{Manghi P}, \textbf{Gandini S},
\textbf{Serrano D}, \textbf{Mizutani S}, \textbf{Shiroma H},
\textbf{Shiba S}, \textbf{Shibata T}, \textbf{Yachida S}, \textbf{Yamada
T}, \textbf{Waldron L}, \textbf{Naccarati A}, \textbf{Segata N},
\textbf{Sinha R}, \textbf{Ulrich CM}, \textbf{Brenner H},
\textbf{Arumugam M}, \textbf{Bork P}, \textbf{Zeller G}. 2019.
Meta-analysis of fecal metagenomes reveals global microbial signatures
that are specific for colorectal cancer. Nature Medicine
\textbf{25}:679.
doi:\href{https://doi.org/10.1038/s41591-019-0406-6}{10.1038/s41591-019-0406-6}.

\hypertarget{ref-vangay_microbiome_2019}{}
23. \textbf{Vangay P}, \textbf{Hillmann BM}, \textbf{Knights D}. 2019.
Microbiome learning repo (ML repo): A public repository of microbiome
regression and classification tasks. Gigascience \textbf{8}.
doi:\href{https://doi.org/10.1093/gigascience/giz042}{10.1093/gigascience/giz042}.

\hypertarget{ref-galkin_human_2018}{}
24. \textbf{Galkin F}, \textbf{Aliper A}, \textbf{Putin E},
\textbf{Kuznetsov I}, \textbf{Gladyshev VN}, \textbf{Zhavoronkov A}.
2018. Human microbiome aging clocks based on deep learning and tandem of
permutation feature importance and accumulated local effects. bioRxiv.
doi:\href{https://doi.org/10.1101/507780}{10.1101/507780}.

\hypertarget{ref-reiman_using_2017}{}
25. \textbf{Reiman D}, \textbf{Metwally A}, \textbf{Dai Y}. 2017. Using
convolutional neural networks to explore the microbiome, pp. 4269--4272.
\emph{In} 2017 39th annual international conference of the IEEE
engineering in medicine and biology society (EMBC).

\hypertarget{ref-fioravanti_phylogenetic_2017}{}
26. \textbf{Fioravanti D}, \textbf{Giarratano Y}, \textbf{Maggio V},
\textbf{Agostinelli C}, \textbf{Chierici M}, \textbf{Jurman G},
\textbf{Furlanello C}. 2017. Phylogenetic convolutional neural networks
in metagenomics. arXiv:170902268 {[}cs, q-bio{]}.
